On centrality of orthogonal $\Kt$

### 2016-05-01 paper.tex
Fixed more comments

\section*{Inroduction}
The main goal of the present paper is to establish Quillen's local-global principle and the centrality of orthogonal $\Kt$ for arbitrary commutative ring $R$.

Recall that to each reduced root system $\Phi$ and a commutative ring $R$ one can associate a split simple group $\GG(\Phi, R)$ called a \emph{Chevalley group}.
For $\Phi$ of rank $\geq 2$ the abstract group $\GG(\Phi, R)$ contains a large normal subgroup $\E(\Phi, R)$ generated by the elementary root unipotents $t_\alpha(\xi)$, $\xi\in R$, $\alpha\in \Phi$.
The quotient group $\Ko(\Phi, R)=\GG(\Phi, R)/\E(\Phi, R)$ is called the \emph{unstable $\Ko$-group}
This group shares many properties with its stable counterpart $\mathrm{SK}_1(R) = \lim\limits_{l\to\infty}\Ko(\rA_\ell, R)$, which, in turn,
is an essential direct summand of the $\Ko$-group ($\Ko(R) = \mathrm{SK}_1(R) \oplus R^*$).

@@

It is classically known that $\Kt(\Phi, R)$ is a central subgroup of $\St(\Phi, R)$ if $R$ is commutative local ring and $\Phi$ has rank $\geq 2$ (see \cite{St73}).
However, to prove that centrality of $\Kt$ holds for \emph{arbitrary} $R$ is much harder.
The first result in this direction was obtained in 1977 by W. van~der~Kallen for $\Phi=\rA_\ell$, $\ell\geq 3$ using the technique called ``another presentation''.
Using a similar approach, the first-named author has established centrality of $\Kt(\rC_\ell, R)$ for $\ell\geq 3$, see \cite{Lav}.
In turn, the key result of this paper is the following theorem.
\begin{tm}\label{tm:centrality}  Let $R$ be an arbitrary commutative ring and $\ell\geq 4$ then 
the group $\Kt(\rD_\ell, R)$ is contained in the centre of $\St(\rD_\ell, R)$. \end{tm}
Notice that the analogous statement in the case $\Phi=\rE_\ell$, $\ell=6,7,8$ has been demonstrated by the second named author in \cite{SCh}.
\cref{tm:centrality} is essentially a corollary of the following local-global principle.
\begin{tm}\label{tm:lg-principle} Let $\Phi$ be an irreducible simply-laced root system of rank $\geq 3$ and $R$ be arbitrary commutative ring.
Then an element $g\in\St(\Phi,\,R[X])$ satisfying $g(0)=1\in\St(\Phi,\,R)$ is trivial in $\St(\Phi,\,R[X])$
if and only if the images $g_M\in\St(\Phi,\,R_M[X])$ of $g$ under localisation morphisms $R\to R_M$ are trivial for all maximal ideals $M\trianglelefteq R$. \end{tm}

@@

Define the map $\psi$ by $\psi(x_{ij}(\xi)) = (X(e_i, (\xi - \pi(\xi)e_j),\ x_{ij}(\pi(\xi)))$. 
We have to check that $\psi$ preserves relations \eqref{add0}--\eqref{ccf2}.
For example, let us check \eqref{ccf2}. 
First of all, recall that in any semidirect product of groups one can compute the commutator of two elements in the following way:
$$[(a, b), (c, d)] = (a \cdot {}^bc \cdot {}^{bdb^{-1}}a^{-1} \cdot {}^{[b, d]}c^{-1} , [b, d]).$$
For $\xi\in R$ set $\xi' = \xi - \pi(\xi)$.
Specializing $a = X(e_i, \xi' e_j)$, $b = x_{ij}(\pi(\xi))$, $c = X(e_j, \eta'e_k)$, $d = x_{jk}(\pi(\eta))$ and using the definition of the conjugation action 
one computes the commutator $[\psi(x_{ij}(\xi), \psi(x_{jk}(\eta)]$ as follows:
$$\left(X(e_i, \xi' e_j) \cdot X(e_j + \pi(\xi)e_i, \eta'e_k) \cdot X(e_i, \pi(\eta)\xi' e_k - \xi' e_j) \cdot X(e_j, -\eta'e_k),\ x_{ik}(\pi(\xi\eta))\right).$$
After applying additivity relations \eqref{add2}, \eqref{add3} and moving the factor $X(e_i, -\xi' e_j)$ to the left hand side of the formula by means of relation \eqref{conj2}
one gets the following expression
$$(X(e_j, \eta'e_k) \cdot X(e_i, (\xi' \eta' + \pi(\xi)\eta' + \pi(\eta)\xi')e_k \cdot X(e_j, -\eta'e_k),\ x_{ik}(\pi(\xi\eta)).$$
Since $\xi'\eta' + \pi(\xi)\eta' + \pi(\eta)\xi' = \xi\eta - \pi(\xi\eta)$ the above expression clearly simplifies to $\psi(x_{ik}(\xi\eta))$, as claimed.
Verification of the fact that $\psi$ preserves relations \eqref{add0}, \eqref{ccf1} is similar to the above computation but is easier.
In particular, one does not need to use relation~\eqref{add3}.

@@

\begin{rk} \label{rk:T123'} Notice that from the proof of the above proposition it follows that one could replace the family of relations \eqref{add3}
in the definition of $\St(n, R, I)$ with the following smaller family:
\setcounter{equation}{2} \renewcommand{\theequation}{T\arabic{equation}'}
\begin{equation} X(e_ir+e_j,\,e_k)=X(e_i,\,rae_k)X(e_j,\,ae_k)\,\text{ for }r\in R,\ a\in I\text{ and } i,j,k\text{ distinct}. \label{add3'} \end{equation} 
This is because only this smaller family of relations is actually used in the construction of $\psi$. \end{rk}

@@

We will prove a somewhat more general result. Let $B$ be a commutative unital ring.
A $B$-module $M$ is called \emph{uniquely $r$-divisible} if for every $m\in M$ there exists a sole $m'\in M$ such that $rm' = m$ (we denote such $m'$ by $\frac{m}{r}$).
Clearly, $M$ is uniquely $r$-divisible if and only if the morphism of principal localisation $\lambda_r\colon M \to M_r$ is an isomorphism.

>>>>

\section*{Introduction}
The main goal of the present paper is to establish Quillen's local-global principle and the centrality of orthogonal $\Kt$ for arbitrary commutative ring $R$.

Recall that to each reduced root system $\Phi$ and a commutative ring $R$ one can associate a split simple group $\GG(\Phi, R)$ called a \emph{Chevalley group}.
For $\Phi$ of rank $\geq 2$ the abstract group $\GG(\Phi, R)$ contains a large normal subgroup $\E(\Phi, R)$ generated by the elementary root unipotents $t_\alpha(\xi)$, $\xi\in R$, $\alpha\in \Phi$, see~\cite{Ta}.
The quotient group $\Ko(\Phi, R)=\GG(\Phi, R)/\E(\Phi, R)$ is called the \emph{unstable $\Ko$-group}
This group shares many properties with its stable counterpart $\mathrm{SK}_1(R) = \lim\limits_{l\to\infty}\Ko(\rA_\ell, R)$, which, in turn,
is an essential direct summand of the $\Ko$-group ($\Ko(R) = \mathrm{SK}_1(R) \oplus R^*$).

@@

It is classically known that $\Kt(\Phi, R)$ is a central subgroup of $\St(\Phi, R)$ if $R$ is a commutative local ring and $\Phi$ has rank $\geq 2$, see \cite{St73}.
We refer to the latter statement as the ``centrality of $\Kt$''.
In~\cite{vdK} using the technique called ``another presentation'' W. van~der~Kallen has shown that the centrality of $\Kt$ holds for $\Phi=\rA_\ell$, $\ell\geq 3$ and arbitrary commutative ring $R$.
Using similar ideas Tulenbaev in~\cite{Tul0} has further generalised van~der~Kallen's result to associative rings that are module-finite over their centres.

Thus, it is natural to expect that the centrality of $\Kt$ holds for arbitrary commutative rings and different types of Chevalley groups, as well.
In \cite{Lav}, \cite{SCh} the authors of the present paper have established the centrality of $\Kt(\rC_\ell, R)$ for $\ell\geq 3$ and $\Kt(\rE_\ell, R)$, $\ell=6,7,8$.
In turn, the key result of this paper is the following theorem.
\begin{tm}\label{tm:centrality}  Let $R$ be an arbitrary commutative ring and $\ell\geq 4$ then 
the group $\Kt(\rD_\ell, R)$ is contained in the centre of $\St(\rD_\ell, R)$. \end{tm}

\cref{tm:centrality} is essentially a corollary of the following local-global principle.
\begin{tm}\label{tm:lg-principle} Let $\Phi$ be an irreducible simply-laced root system of rank $\geq 3$ and $R$ be arbitrary commutative ring.
Then an element $g\in\St(\Phi,\,R[X])$ satisfying $g(0)=1\in\St(\Phi,\,R)$ is trivial in $\St(\Phi,\,R[X])$
if and only if the images $g_M\in\St(\Phi,\,R_M[X])$ of $g$ under localisation morphisms $R\to R_M$ are trivial for all maximal ideals $M\trianglelefteq R$. \end{tm}

@@

Define the map $\psi$ by $\psi(x_{ij}(\xi)) = (X(e_i, (\xi - \pi(\xi)e_j),\ x_{ij}(\pi(\xi)))$. 
We have to check that $\psi$ preserves relations \eqref{add0}--\eqref{ccf2}.
For example, let us check \eqref{ccf2}. 
First of all, recall that in any semidirect product of groups one can compute the commutator of two elements in the following way:
$$[(a, b), (c, d)] = (a \cdot {}^bc \cdot {}^{bdb^{-1}}a^{-1} \cdot {}^{[b, d]}c^{-1} , [b, d]).$$
For $\xi\in R$ set $\xi' = \xi - \pi(\xi)$.
Specializing $a = X(e_i, \xi' e_j)$, $b = x_{ij}(\pi(\xi))$, $c = X(e_j, \eta'e_k)$, $d = x_{jk}(\pi(\eta))$ and using the definition of the conjugation action 
one computes the commutator $[\psi(x_{ij}(\xi), \psi(x_{jk}(\eta)]$ as follows:
\setcounter{equation}{2}
\begin{equation} \left(X(e_i, \xi' e_j) \cdot X(e_j + \pi(\xi)e_i, \eta'e_k) \cdot X(e_i, \pi(\eta)\xi' e_k - \xi' e_j) \cdot X(e_j, -\eta'e_k),\ x_{ik}(\pi(\xi\eta))\right). \label{eq:proofS3}\end{equation}
\setcounter{lm}{3}
After applying additivity relations \eqref{add2}, \eqref{add3} and moving the factor $X(e_i, -\xi' e_j)$ to the left hand side of the formula by means of relation \eqref{conj2}
one gets the following expression
$$(X(e_j, \eta'e_k) \cdot X(e_i, (\xi' \eta' + \pi(\xi)\eta' + \pi(\eta)\xi')e_k \cdot X(e_j, -\eta'e_k),\ x_{ik}(\pi(\xi\eta)).$$
Since $\xi'\eta' + \pi(\xi)\eta' + \pi(\eta)\xi' = \xi\eta - \pi(\xi\eta)$ the above expression clearly simplifies to $\psi(x_{ik}(\xi\eta))$, as claimed.
Verification of the fact that $\psi$ preserves relations \eqref{add0}, \eqref{ccf1} is similar to the above computation but is easier.
In particular, one does not need to use relation~\eqref{add3}.

@@

\begin{rk} \label{rk:T123'} Notice that from the proof of the above proposition it follows that one could replace the family of relations \eqref{add3}
in the definition of $\St(n, R, I)$ with the following smaller family:
\setcounter{equation}{2} \renewcommand{\theequation}{T\arabic{equation}'}
\begin{equation} X(e_ir+e_j,\,e_k)=X(e_i,\,rae_k)X(e_j,\,ae_k)\,\text{ for }r\in R,\ a\in I\text{ and } i,j,k\text{ distinct}. \label{add3'} \end{equation} 
This follows from the fact that only this smaller family of relations was actually used in the construction of the map $\psi$
(more precisely, it was used to simplify the second factor of~\eqref{eq:proofS3}). \end{rk}

@@

We will prove a somewhat more general result. Let $B$ be a commutative unital ring.
We call an ideal $I$ of $B$ \emph{uniquely $r$-divisible} if for every $m\in I$ there exists only one $m'\in I$ such that $rm' = m$ (we denote such $m'$ by $\frac{m}{r}$).
Clearly, $I$ is uniquely $r$-divisible if and only if the restriction of the morphism of principal localisation $\lambda_r\colon R \to R_r$ to $I$ is an isomorphism.
